Measles risk assessment and cost interim report\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-07-02 interimreport1.Rnw
Summary data, figures, and revised report code
Include summary data and figures in PDF form to work into the report
without having to reload the data.

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}

@@

@@

A previously-published model of the dynamics of measles infections
in New Zealand is used to evaluate the present vaccination strategy of
MMR1 at 15 months and MMR2 before 5 years. The results show that
achieving coverage of 90\% or better at both vaccination opportunities is
necessary if future epidemics of measles are to be prevented.

@@

A variety of similar models for measles vaccination strategies have been developed
by other authors for various regions (Agur et al 1993; Babad et al 1995;
Edmunds et al 2000; Gay et al 1998; Wallinga et al 2001). These have invariably
been based on sets of nonlinear differential equations, and the conclusions
reached have been similar. The differences in the models have been in the details
of the representation of the infectious period, and in the ways in which
the age and contact structures of the population have been specified. As the
time-course of a measles epidemic is short compared to that of demographic
changes, details of the epidemic process are not relevant when examining population
level changes over many years. In fact, it has been shown (Diekmann
& Heesterbeek 2000) that the number infected in an epidemic depends only
on the basic reproduction number (see below), and the numbers that are initially
susceptible in each sub-population. It is therefore important to have a
realistic representation of the population contact structure. The Royal Society
(UK) report on the 2001 epidemic of Foot and Mouth Disease advocated that
contact structures for infectious diseases should be described before epidemics
occur, so that models may be speedily prepared. The report presented the
New Zealand measles model as an example of the successful use of modelling
in epidemiology (Anon. 2002b).

@@

In order to assess the efficacy of the vaccination campaign in the absence of
precise information on levels of vaccination coverage, we assumed a range of
plausible coverages at MMR1 and MMR2 and determined the corresponding
value of Rv for each. For selected combinations we also numerically solved
the model over a period of 150 years, not as a prediction of what will happen
over that time period but in order to determine what the expected frequency
of epidemics would be if this level of vaccination were to be in maintained. In
all cases, if Rv < 1 epidemics are prevented.

Method
For this exercise we modelled the New Zealand population in the same way
as in Roberts & Tobias (2000). We approximated the population structure by
assuming that the birth rate was constant, and that deaths before the age of
25 could be neglected. Hence the four age groups active in the epidemic have
constant size. The age groups were taken as 6 - 15 months, 15 months - 5 years,
5 years - 11 years and 11 years - 25 years. Hence MMR1 is applied as children
pass from the first to the second group, and MMR2 from the second to the
third. The same contact matrix was used as in Roberts & Tobias (2000). For
a range of vaccination coverages for MMR1 and MMR2 (see Table 1) the basic
reproduction number of the infection under vaccination, Rv, was calculated.
In addition, for each combination of coverages the differential equation model
was solved numerically until the epidemics settled into a regular pattern, and
the inter-epidemic period was determined.

@@

Table 1: The basic reproduction number of the infection under vaccination,
Rv, for various coverage levels of MMR1 and MMR2. Superscripts refer to
the figure number where numerical solutions are presented. If Rv < 1 then
epidemics will not occur.

When no vaccination is applied, the system settles down to an epidemic every
two years. This corresponds to an estimated R0 = 12.5, similar to that which
was observed in New Zealand in the 1960s (see Fig.1). The minimum level of
vaccination investigated (MMR1 = 80\%, MMR2 = 60\%) resulted in Rv = 1.72
and a pattern of epidemics settling down to one every 12 years (see Fig.2).
The other combinations of MMR1 and MMR2 coverage levels that resulted
in Rv > 1 all resulted in epidemics recurring at intervals of between 10 and
20 years (Figs. 3-7). All other combinations of MMR1 and MMR2 resulted
in Rv > 1 and the system settling down to a steady uninfected state with no
epidemics, see Table 1 and Fig.8 for an example.

Figures 1 & 2. The epidemic pattern with no vaccination (top) and with
MMR1 = 80\%, MMR2 = 60\% (bottom). The continuous curves show the
cumulative numbers of susceptibles in the population in the age groups 6
months-15 months, 15 months - 5 years, 5 years - 11 years and 11 years -25
years, as determined by the model. The spikes show the numbers infectious
during epidemics (×10).

Figures 3 & 4. The epidemic pattern with with MMR1 = 80\%, MMR2 = 70\%
(top) and with MMR1 = 80\%, MMR2 = 80\% (bottom). The continuous
curves show the cumulative numbers of susceptibles in the population in the
age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
11 years -25 years, as determined by the model. The spikes show the numbers
infectious during epidemics (×10).

Figures 5 & 6. The epidemic pattern with MMR1 = 85\%, MMR2 = 60\%
(top) and with MMR1 = 85\%, MMR2 = 70\% (bottom). The continuous
curves show the cumulative numbers of susceptibles in the population in the
age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
11 years -25 years, as determined by the model. The spikes show the numbers
infectious during epidemics (×10).

Figures 7 & 8. The epidemic pattern with MMR1 = 90\%, MMR2 = 60\%
(top) and with MMR1 = 80\%, MMR2 = 85\% (bottom). The continuous
curves show the cumulative numbers of susceptibles in the population in the
age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
11 years -25 years, as determined by the model. The spikes show the numbers
infectious during epidemics (×10).

The model developed by Roberts & Tobias (2000) supported the change in the
immunisation schedule that took effect in January 2001, at which time MMR2
was changed from delivery at 11 years to delivery before the age of five. These
results were in line with those obtained by other authors, for example: Babad
et al (1995) advocated a two-dose schedule for England and Wales, with the
second vaccination given at age four; and Gay et al (1998) recommended a
second vaccination at either 18 months or five years, to complement the first
vaccination at 12 months in Canada. In addition, Agur et al (1993) found that
vaccinating 85\% of susceptible children aged one to seven years at five-yearly
intervals would prevent epidemics in Israel. These authors all agree that two
vaccinations at no less than five years apart are necessary to prevent measles
epidemics.
A different approach was taken by Wallinga et al (2001). These authors took
existing policies in eight European countries and estimated the coverage rates
required to reduce Rv below one. They found that results depended on the
age at delivery, but no strategy succeeded if coverage rates were below approximately
87\%. Our results presented in Table 1 are similar where coverage is
assumed to be the same at MMR1 and MMR2. The results also show the absolute
necessity of maintaining high coverage rates in order to prevent future
epidemics. It is difficult to estimate the proportion of the school-age population
that have been effectively immunised, but this is continuously being
diluted by children who are not immunised. Wallinga et al (2000) noted that
in Italy only MMR1 was offered at 18 months with no second vaccination opportunity,
and that under most plausible assumptions for contact rates even
100\% coverage would be insufficient to prevent epidemics.
A mathematical model for measles vaccination 11
Our final conclusion must depend on how much faith is placed in the model.
The results are consistent with those of other authors, and from Table 1 it
could be deduced that 85\% coverage at MMR1 and MMR2 could be sufficient
to prevent future measles epidemics. However, a study by Glass et al (2004)
in the Netherlands showed that high overall levels of measles vaccination can
obscure pockets of poor coverage, resulting in localised regions with increased
risk of infection. Our results indicate that overall targets or 90% or more will
need to be achieved to prevent future epidemics in New Zealand.

@@

>>>>

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}

@@

\section{Future cost-benefit analyses}
Using the results above we aim to:
\begin{itemize}
\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date and in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified if the multivariate analyses can be performed once appropriate denominator data are obtained.
\end{itemize}

@@

A previously-published model of the dynamics of measles infections
in New Zealand has been used to evaluate the vaccination strategy in New Zealand of
MMR1 at 15 months and MMR2 before 5 years. The results show that
achieving coverage of >90\% at both vaccination opportunities is
necessary if future epidemics of measles are to be prevented. [AND LIKELY MORE?!]

@@

A variety of similar models for measles vaccination strategies have been developed
by other authors for various regions (Agur et al 1993; Babad et al 1995;
Edmunds et al 2000; Gay et al 1998; Wallinga et al 2001). These have invariably
been based on sets of nonlinear differential equations, and the conclusions
reached have been similar. The differences in the models have been in the details
of the representation of the infectious period, and in the ways in which
the age and contact structures of the population have been specified.

The model developed by Roberts & Tobias (2000) supported the change in the
immunisation schedule that took effect in January 2001, at which time MMR2
was changed from delivery at 11 years to delivery before the age of five. These
results were in line with those obtained by other authors, for example: Babad
et al (1995) advocated a two-dose schedule for England and Wales, with the
second vaccination given at age four; and Gay et al (1998) recommended a
second vaccination at either 18 months or five years, to complement the first
vaccination at 12 months in Canada. In addition, Agur et al (1993) found that
vaccinating 85\% of susceptible children aged one to seven years at five-yearly
intervals would prevent epidemics in Israel. These authors all agree that two
vaccinations at no less than five years apart are necessary to prevent measles
epidemics.
A different approach was taken by Wallinga et al (2001). These authors took
existing policies in eight European countries and estimated the coverage rates
required to reduce Rv below one. They found that results depended on the
age at delivery, but no strategy succeeded if coverage rates were below approximately
87\%. Our results presented in Table 1 are similar where coverage is
assumed to be the same at MMR1 and MMR2. The results also show the absolute
necessity of maintaining high coverage rates in order to prevent future
epidemics. It is difficult to estimate the proportion of the school-age population
that have been effectively immunised, but this is continuously being
diluted by children who are not immunised. Wallinga et al (2000) noted that
in Italy only MMR1 was offered at 18 months with no second vaccination opportunity,
and that under most plausible assumptions for contact rates even
100\% coverage would be insufficient to prevent epidemics.
Our final conclusion must depend on how much faith is placed in the model.
The results are consistent with those of other authors, and from Table 1 it
could be deduced that 85\% coverage at MMR1 and MMR2 could be sufficient
to prevent future measles epidemics. However, a study by Glass et al (2004)
in the Netherlands showed that high overall levels of measles vaccination can
obscure pockets of poor coverage, resulting in localised regions with increased
risk of infection. Our results indicate that overall targets or 90\% or more will
need to be achieved to prevent future epidemics in New Zealand.

@@

@@

\begin{figure}[p]
     \centering
     \includegraphics[width=0.8\textwidth]{averageR0.pdf}
     \caption{Test image}
     \label{fig:test image}
\end{figure}

@@

\section{Future cost-benefit analyses}
Using the results above we aim to:
\begin{itemize}
\item Update previous ODE models of measles outbreak risk in the overall population according the differing vaccine coverage scenarios.
\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
\end{itemize}
